---
title: "GBM Patient Journey"
author: "Blake Easton"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(Hmisc)
library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
library(dplyr)
library(magrittr)
library(cluster)
library(factoextra)
knitr::opts_chunk$set(echo = TRUE)
```

## Data Prep

```{r data-prep}
path <- "variable-dosing_7119.xlsx"
studies <- read_excel(path, sheet="Studies")
arms <- read_excel(path, sheet="Arms")
dropout <- read_excel(path, sheet="Dropout")

studies <- studies %>%
  mutate(NCT = as.factor(NCT))

arms <- arms %>%
  mutate(NCT = as.factor(NCT)
        ,Arm = as.factor(Arm)
        )

dropout <- dropout %>%
  mutate(NCT = as.factor(NCT)
        ,Arm = as.factor(Arm)
        ,Reason = as.factor(Reason)
        )

# Here's a more uniform & reliable way to calculate dropout rates by arm
dropout <- dropout %>%
  mutate(burden_related = Reason %in% c("Withdrawal by Subject"
                                       ,"Patient Refusal"
                                       ,"Study Compliance")) %>%
  group_by(NCT,Arm) %>%
  summarize(brDropout = sum(Dropouts * burden_related)) # NB: logical multiplies like {0,1}

# At this point, 'dropout' has 46 rows -- one for each arm.
# We can then join the arm enrollments, and calculate brDropoutPercent:
dropout <- dropout %>%
  left_join(arms[,c("NCT","Arm","Enrollment","Burden","BurdenTwo")], by=c("NCT","Arm")) %>%
  mutate(brDropoutRate = brDropout / Enrollment)

# We may as well unify 'dropout' with 'arms' at this point, for simplicity:
arms <- arms %>% left_join(dropout[,c("NCT","Arm","brDropout","brDropoutRate")]
                           , by=c("NCT","Arm"))

#build dropout rate col
arms$brDropoutLabel <- ifelse(arms$brDropoutRate == 0, "LowDrop",
                              ifelse(arms$brDropoutRate < 0.07692307, "MedDrop", "HighDrop"))

```

## Integrity Checks
```{r integrity-checks}
stopifnot(nlevels(dropout$Arm) == nlevels(arms$Arm))
duped <- dropout %>%
  group_by(Arm) %>%
  summarize(duped = length(unique(NCT)) > 1)
arm_names_unique <- all(!duped$duped)
stopifnot(all(!duped$duped))
```

## Cluster Analysis

```{r patient-burden, fig.height=5, fig.cap="TODO"}
#create data frame for kmeans
df_arm <- data.frame(select(arms, `Universal IV`, Radiation, Surgery, SOC, Recurrent))
rownames(df_arm) <- arms$Arm

#build cluster graph
set.seed(2019) # ensure reproducible cluster numbering
k2 <- kmeans(scale(df_arm), centers = 3, nstart=25, algorithm ="Forgy")
fviz_cluster(k2, data=df_arm, repel=TRUE, labelsize = 7)

# #utilize k2 data / format
# NB: k2$cluster is now a *named* integer vector
#     with names perfectly aligned with arms$Arm,
#     so that we may easily attach it to 'arms':
stopifnot(all(arms$Arm == names(k2$cluster)))
arms$cluster <- k2$cluster

# #build boxplot
bwplot(cluster ~ (100*brDropoutRate), data=arms)

```
## A deep dive into the cluster analysis

The `k2` cluster object is a list, with a number of components. The first of these, `k2$centers`, helps us to appreciate how the clustering algorithm has partitioned your data:

```{r cluster-constrasts, echo=TRUE, results='verbatim'}
ctrs <- round(k2$centers, 1) # rounding avoids distraction by excessive detail
rownames(ctrs) <- paste("Cluster", 1:3)
kable(ctrs) 
```


From the columns of this table, we can see that Surgery is the distinctive feature of Cluster 2, while Cluster 1 is marked by absence of both Radiation and SOC, and the presence of Recurrent disease. (Specifically, observe that all three column vectors are close to $\pm(-1,1,1)$.)

This suggests that these three attributes do not contribute independent information that discriminates between the arms. Why this is so may be appreciated by noting these almost perfectly diagonal cross-tabulations:

```{r Rads_SOC}
xtabs(~ SOC + Radiation, data=arms)
```

```{r Recurrent_SOC}
xtabs(~ SOC + !Recurrent, data=arms)
```

```{r Rads_Recurrent}
xtabs(~ Radiation + !Recurrent, data=arms)
```

Another way to appreciate these facts is by cross-tabulating the cluster designations against each of these facts independently:

```{r cluster_Recurrent}
xtabs(~ k2$cluster + arms$Recurrent)
```

```{r cluster_Radiation}
xtabs(~ k2$cluster + arms$Radiation)
```

```{r cluster_SOC}
xtabs(~ k2$cluster + arms$SOC)
```

Finally, we ought to note how the `Universal IV` factor cross-tabulates against these clusters:

```{r cluster_UnivIV}
xtabs(~ k2$cluster + arms$`Universal IV`)
```

This crosstab helps explain the column vector $(0.0, -0.9, 0.1)$ for `Universal IV` in the `k2$centers` table, as reflecting the approximate balance on this variable within each of clusters 1 and 3.

## Sample correspondence analysis

Might some of these close factor alignments have emerged from a multiple correspondence analysis?

```{r MCA-prep}
# Prepare data frame for MCA
mca_df <- arms %>%
  mutate(
    Recur = factor(Recurrent, levels=c(T,F), labels=c("Recurrent","NewlyDxed"))
  , Burden = ordered(Burden, levels=c("Low","Medium","High"))
  , UnivIV = factor(`Universal IV`, levels=c(T,F), labels=c("IV","Oral"))
  , SOC = factor(SOC, levels=c(T,F), labels=c("SOC","notSOC"))
  , Rads = factor(Radiation, levels=c(T,F), labels=c("Rads","noRads"))
  , Dropout = ordered(brDropoutLabel, levels=c("LowDrop", "MedDrop", "HighDrop"))
  )
```

```{r MCA, fig.cap="Multiple correspondence analysis (MCA) for several key design factors of the study arms. The close clustering of radiation, standard-of-care (SOC) and newly-diagnosed indicates that these design features strongly correlate, and likewise with the complementary features. Interestingly, the Oral-IV dichotomy seems orthogonal to these other features. Oral medication and low burden (as graded) appear related."}

library(ca)
mca <- mjca(mca_df[,c("Recur","Burden","UnivIV","SOC","Rads","Dropout")])
plot(mca, collabels="level", map="rowgreen", mass=TRUE)
```

## Patient Burden (Emily)

```{r patient-burden-emily, fig.height=5, fig.cap="**TODO**"}
#create data frames for combination

#build chart
bxp <- boxplot(100*brDropoutRate ~ Burden, data=arms,
        horizontal = TRUE,
        xlab = "% of Patient Dropouts",
        col = "grey"
        )

stripchart(brDropoutRate ~ Burden, data=arms,
          add = TRUE,
          pch = 20,
          method = "jitter",
          col = "blue"
          )

#format text
low <- format(round(bxp$stats[3,1], 2), nsmall = 2)
medium <- format(round(bxp$stats[3,2], 2), nsmall = 2)
high <- format(round(bxp$stats[3,3], 2), nsmall = 2)

#add text
text(x=0.15, y=1, labels=low)
text(x=0.315, y=2, labels=medium)
text(0.30, y=3, labels=high)
title("Patient Dropout % by Burden Level")

```



